PMV formulations: which one is the most accurate?

### 2023-04-18 Manuscript/src/sections/introduction.tex
feat: reviewed methodology

Some notable works include, the work from \mycite{Doherty1988} who evaluated the ability of the \ac{pmv} and \ac{set} models to predict several physiological variables (i.e., skin temperature, core temperature, and skin wettedness) under a wide range of still air environments and metabolic rates.
They concluded that the \ac{pmv} model is accurate for simulations of resting subjects but its accuracy decreases as a function of metabolic rate.
Humphreys and Nicol estimated the effects of measurement and formulation error on predicting thermal sensation using the \ac{pmv}~\cite{Humphreys2000}.
They used the ASHRAE Thermal Comfort database v1.0 and determined that the measurement error and the error introduced by the \ac{pmv} model formulation had a similar and non-negligible contribution.
Humphreys and Nicol also determined the validity of the \ac{pmv} for predicting comfort votes collected in field studies~\cite{Humphreys2002}.
They concluded that the \ac{pmv} range of applicability should be significantly reduced and it fails to predict the extent of thermal dissatisfaction of people in buildings.
The \ac{pmv} was free from bias only when it was used to predict thermal neutrality.
While, \mycite{Cheung2019} determined the accuracy of the \ac{pmv} model, by comparing its results with the \ac{tsv} from the ASHRAE Global Thermal Comfort Database II.
Finally, while \mycite{Yao2022} compare the \ac{pmv} and \ac{pmv-ce} models, their aim was primarily to compare these two formulations with other adaptive \ac{pmv} formulations, hence, they do not provide a detailed analysis on the prediction accuracy of the two models.
In particular, they focus significantly how the models perform in different climates and when applied to people from different world regions, but for example they fail to report how the \ac{pmv} accuracy varies as a function of \ac{tsv} and how the bias of the model varies as a function of different environmental parameters.

>>>>

Some notable works include, the work from \mycite{Doherty1988} who evaluated the ability of the \ac{pmv} and \ac{set} models to predict several physiological variables (i.e., skin temperature, core temperature, and skin wettedness) under a wide range of still air environments and metabolic rates.
They concluded that the \ac{pmv} model is accurate for simulations of resting subjects but its accuracy decreases as a function of metabolic rate.
Humphreys and Nicol estimated the effects of measurement and formulation error on predicting thermal sensation using the \ac{pmv}~\cite{Humphreys2000}.
They used the ASHRAE Thermal Comfort database v1.0 and determined that the measurement error and the error introduced by the \ac{pmv} model formulation had a similar and non-negligible contribution.
Humphreys and Nicol also determined the validity of the \ac{pmv} for predicting comfort votes collected in field studies~\cite{Humphreys2002}.
They concluded that the \ac{pmv} range of applicability should be significantly reduced and it fails to predict the extent of thermal dissatisfaction of people in buildings.
The \ac{pmv} was free from bias only when it was used to predict thermal neutrality.
While, \mycite{Cheung2019} determined the accuracy of the \ac{pmv} model, by comparing its results with the \ac{tsv} from the ASHRAE Global Thermal Comfort Database II.
Finally, while \mycite{Yao2022} compare the \ac{pmv} and \ac{pmv-ce} models, their aim was primarily to compare these two formulations with other adaptive \ac{pmv} formulations, hence, they do not provide a detailed analysis on the prediction accuracy of the two models.
In particular, they focus significantly how the models perform in different climates and when applied to people from different world regions, but for example they fail to report how the \ac{pmv} accuracy varies as a function of \ac{tsv}.
Moreover, their analysis only reported the mean bias of the different models.
This as depicted by \mycite{Humphreys2000} does not provide sufficient insights and information in depicting which model is more accurate in predicting \ac{tsv} particularly because the ASHRAE Global Thermal Comfort Database II v2.1 is an unbalanced dataset.

### 2023-04-18 Manuscript/src/sections/methodology.tex
feat: reviewed methodology

The \gls{db2} has \var{entries_db_all} independent recordings from various field investigations throughout the world and it is the largest of its kind.
\Ac{tsv} were collected using a \gls{rhrn} survey and environmental parameters were measured and logged while participants completed their survey.
The Comfort DB contains data collected by researchers in cross-sectional field studies that have been published in peer reviewed journals.
% We compared the results obtained by the \ac{pmv} and the \gls{pmv-ce} with the self-reported \ac{tsv}.
%    All the results were calculated using the function pmv_ppd [6] included in pythermalcomfort v1.10.0 a Python package for thermal comfort calculations.

\subsection{Data Preparation and Cleaning}\label{subsec:data-processing-and-cleaning}
Not all researchers logged all the six parameters required to calculate the PMV.\@
We, therefore, decided to remove all entries that did not have any of the following variables: \ac{tdb}, \ac{rh}, \ac{v}, \ac{clo}, \ac{met}, and \ac{tsv}.
These constraints filtered out approximately \var{entries_db_valid} of the data in the Comfort DB.\@
We decided to keep entries that did not have \ac{tr}.
Since \ac{tr} is a required input in the \ac{pmv} model, it was calculated using the following equation: 2\ac{to} – \ac{tdb}.
Alternatively, if \ac{to} was not available we assumed \ac{tr} equal to \ac{tdb}.
We are aware that these assumptions may introduce an error, but removing samples without \ac{tr} would cause an additional \var{entries_db_valid_no_tr} drop in the data used to compare the models.
There is also evidence that in many conditions is possible to assume \ac{tr} equal to \ac{tdb} without committing a major error~\cite{Dawe2020}.

Both the \gls{55} and \gls{7730} specify a set of applicability limits, we hence filtered out those data points which did not meet the inclusion criteria of both Standards.
The rationale is that the model accuracy should only be tested in within its applicability limits.
The limits we used are as follows: \qty{10}{\celsius} $\leq$ \ac{tdb} $\leq$ \qty{30}{\celsius}, \qty{10}{\celsius} $\leq$ \ac{tr} $\leq$ \qty{40}{\celsius}, \qty{0}{\m\per\s} $\leq$ \ac{v} $\leq$ \qty{1}{\m\per\s}, \qty{0}{clo} $\leq$ \ac{clo} $\leq$ \qty{1.5}{clo}, 0 Pa $\leq$ water vapour partial pressure $\leq$ 2000 Pa, and \qty{1}{met} $\leq$ \ac{met} $\leq$ \qty{4}{met}.
We then calculated the adjusted total clothing insulation and relative airspeed as required by both Standards.
We used these inputs to calculate the \ac{pmv} values.  % todo check with marcel if this is required in his model too

@@

The values of \ac{pmv} and \gls{pmv-ce} are available inside the \gls{db2}.
We calculated the values of \gls{pmvs}, \gls{pmvg}, and \ac{athb} using pythermalcomfort v2.5.1~\cite{Tartarini2020a}.
We used Python to analyse and plot the results.

We are committed to reproducible research hence we have shared the source code and the dataset we used publicly at this URL: \textbf{provide URL} so other users can test different assumptions.

\subsection{Model Validation}\label{subsec:model-validation}
The \ac{pmv} model was developed to predict the average thermal sensation of a large group of occupants sharing the same environment.
While the \gls{db2} contains data from a sufficiently wide population no information is available on where exactly these data were collected.
We only have access to building level data and no information about weather a group of occupants was sharing the same thermal environment.
We, therefore, decided to first group the participants responses by their \ac{tsv} and we determined how many of these votes were correctly labelled by the \ac{pmv} model.
This approach is useful to perform an overall assessment of the accuracy of the model, but it introduces an error due to the rounding of the \ac{pmv} models output and \ac{tsv} values.
The latter were sometimes collected using a continuous scale and not a discrete one, in other words they could take any value in the range of \numrange{-3.0}{3.0}.
To remove this rounding error, we then went on subtracted the \ac{tsv} value from the respective \ac{pmv} value.
These deltas quantify the success of the model in predicting \ac{tsv}.
However, on their own are a low precision estimate of the overall accuracy of the model~\cite{Humphreys2002}.
These values were then binned using several independent variables (e.g., \ac{tdb}, \ac{rh}).
If the \ac{pmv} formulation is bias-free, the distribution of any batch derived from these differences would have a mean value that was little different from zero.
The standard deviation would reflect the combined effect of the people's individual differences as well as any errors in the \ac{pmv} formulation~\cite{Humphreys2002}.
According to \mycite{Humphreys2002} the model can be considered accurate if the above-mentioned deltas are within \numrange{-0.25}{0.25}.
% todo why not 0.5 since the delta between a PMV of 2.5 and a self-reported vote of either 2 or 3 (closest points to 2.5) is 0.5?

\subsection{Sources of Error}\label{subsec:sources-of-error}
One minor but not negligible source of error is the use of a discrete scale to assess thermal sensation.
Most of the \ac{tsv} in the \gls{db2} were collected using a discrete scale while the \ac{pmv} output is continuous.
For example, even if the model was 100~\% accurate and the estimated \ac{pmv} values was \num{2.5} the person could only report to be `warm (+2)' or `hot (+3)'.

Measurement errors are also inevitably present in the dataset.
These will add a random error to the model output.

Errors are also caused by the fact that the \ac{pmv} model is an approximation of a highly complex system (the human body and mind) which interacts with the surrounding environment.
Fanger could not, therefore, include all the factors that contribute to the comfort vote and some of them have been omitted.
Finally, it should be noted that the \ac{pmv} model has been developed considering based on the assumption of steady-state heat transfer, however, this never precisely occurs since the human thermo regulatory system is always actively engaged to ensure a stable core temperature.
However, as \mycite{Humphreys2002} note, these approximations and errors present in the \ac{pmv} model, if the model is accurate, should combine to produce a quasi-random error in the prediction.
This would increase the standard deviation of the delta between the \ac{tsv} and \ac{pmv}, but if the model is sound, they should produce a negligible bias~\cite{Humphreys2002}.
On the other hand, if the model is neither accurate nor representative of the real system a significant bias will arise.

>>>>

The \gls{db2} has \var{entries_db_all} independent recordings from various field investigations throughout the world, and it is the largest of its kind~\cite{FoldvaryLicina2018, db2dryad}.
\Ac{tsv} were collected using a \gls{rhrn} survey and environmental parameters were measured and logged in the proximity of the participants when they completed the survey.
The Comfort DB contains data collected by researchers in cross-sectional field studies that have been published in peer reviewed journals.
 We compared the results obtained by the \ac{pmv} and the \gls{pmv-ce} with the self-reported \ac{tsv}.
Results were calculated using the function `pythermalcomfort.models.pmv\_ppd' included in pythermalcomfort v2.7.0 a Python package for thermal comfort calculations~\cite{Tartarini2020a}.

\subsection{Data Preparation and Cleaning}\label{subsec:data-processing-and-cleaning}
Not all researchers logged all the six parameters required to calculate the PMV.\@
We, therefore, decided to remove all entries that did not have any of the following variables: \ac{tdb}, \ac{rh}, \ac{v}, \ac{clo}, \ac{met}, and \ac{tsv}.
These constraints filtered out approximately \var{entries_db_valid} of the data in the Comfort DB.\@
We decided to keep entries that did not have \ac{tr}.
Since \ac{tr} is a required input in the \ac{pmv} model, it was calculated using the following equation: $2 t_{o} – t_{db}$.
Alternatively, if \ac{to} was not available we assumed \ac{tr} equal to \ac{tdb}.
We are aware that these assumptions may introduce an error, but removing samples without \ac{tr} would cause an additional \var{entries_db_valid_no_tr} drop in the data used to compare the models.
There is also evidence that in many conditions is possible to assume \ac{tr} equal to \ac{tdb} without committing a major error~\cite{Dawe2020}.

Both the \gls{55} and \gls{7730} specify a set of applicability limits, we hence filtered out those data points which did not meet the inclusion criteria of both Standards.
The rationale is that the model accuracy should only be tested in within its applicability limits.
The limits we used are as follows:
\qty{10}{\celsius} $\leq$ \ac{tdb} $\leq$ \qty{30}{\celsius},
\qty{10}{\celsius} $\leq$ \ac{tr} $\leq$ \qty{40}{\celsius},
\qty{0}{\m\per\s} $\leq$ \ac{v} $\leq$ \qty{1}{\m\per\s},
\qty{0}{clo} $\leq$ \ac{clo} $\leq$ \qty{1.5}{clo},
0 Pa $\leq$ water vapour partial pressure $\leq$ 2000 Pa,
and \qty{1}{met} $\leq$ \ac{met} $\leq$ \qty{4}{met}.
We then calculated the adjusted total clothing insulation and relative airspeed as required by both Standards.
We used these inputs to calculate the \ac{pmv} values.

@@

We used Python to analyse and plot the results.

We are committed to reproducible research hence we have shared the source code and the dataset we used publicly at this URL: \textbf{provide URL} so other users can test different assumptions.  % todo add URL

\subsection{Model Validation}\label{subsec:model-validation}
The \ac{pmv} model was developed to predict the average thermal sensation of a large group of occupants sharing the same environment.
While the \gls{db2} contains data from a sufficiently wide population no information is available on where exactly these data were collected.
We only have access to building level data and no information about weather a group of occupants was sharing the same thermal environment.
We, therefore, decided to first group the participants responses by their \ac{tsv} and we determined how many of these votes were correctly labelled by the \ac{pmv} model.
This approach is useful to perform an overall assessment of the accuracy of the model, but it introduces an error due to the rounding of the \ac{pmv} models output and \ac{tsv} values.
The latter were sometimes collected using a continuous scale and not a discrete one, in other words they could take any value in the range of \numrange{-3.0}{3.0}.
To remove this rounding error, we then went on subtracted the \ac{tsv} value from the respective \ac{pmv} value.
These deltas quantify the success of the model in predicting \ac{tsv}.
However, on their own are a low precision estimate of the overall accuracy of the model~\cite{Humphreys2002}.
These values were then binned using several independent variables (e.g., \ac{tdb}, \ac{rh}).
If the \ac{pmv} formulation is bias-free, the distribution of any batch derived from these differences would have a mean value that was little different from zero.
The standard deviation would reflect the combined effect of the people's individual differences as well as any errors in the \ac{pmv} formulation~\cite{Humphreys2002}.
According to \mycite{Humphreys2002} the model can be considered accurate if the above-mentioned deltas are within \numrange{-0.25}{0.25}.
% todo why not 0.5 since the delta between a PMV of 2.5 and a self-reported vote of either 2 or 3 (the closest points to 2.5) is 0.5?

\subsection{Performance Metrics}\label{subsec:performance-metrics}
For classification problems that involve unbalanced dataset the simple accuracy is a misleading metric to use to quantify the performance of a predictive model~\cite{Chawla2005}.
We, therefore, decided to use the F1 score when determining the performance of the \ac{pmv} in predicting the \ac{tsv} reported by a participant.
The F1 score can be interpreted as a harmonic mean of the precision and recall, where an F1 score reaches its best value at 1 and worst score at 0.
Precision (also called positive predictive value) is the fraction of relevant instances among the retrieved instances, while recall (also known as sensitivity) is the fraction of relevant instances that were retrieved.
The F1-micro score calculate metrics globally by counting the total true positives, false negatives and false positives.
The F1-macro calculate metrics for each label, and find their unweighted mean.
The F1-weighted calculate metrics for each label, and find their average weighted by support (the number of true instances for each label).
This alters ‘macro’ to account for label imbalance; it can result in an F-score that is not between precision and recall.
% todo descibe also the other metrics we used like R2

### 2023-04-18 Manuscript/src/sections/titleAbstract.tex
feat: reviewed methodology

    \author[label1]{Federico Tartarini\corref{mycorrespondingauthor}}
    \ead{federicotartarini@berkeley.edu}
    \author[label2]{Stefano Schiavon}

    \address[label1]{Berkeley Education Alliance for Research in Singapore, Singapore}
    \address[label2]{Center for the Built Environment, University of California, Berkeley, CA, USA}

>>>>

    \author[label1,label2]{Federico Tartarini\corref{mycorrespondingauthor}}
    \ead{federicotartarini@berkeley.edu}
    \author[label3]{Stefano Schiavon}

    \address[label1]{Berkeley Education Alliance for Research in Singapore, Singapore}
    \address[label2]{Heat and Health Research Incubator, Faculty of Health and Medicine, University of Sydney, Sydney, AU}
    \address[label3]{Center for the Built Environment, University of California, Berkeley, CA, USA}
